Dependent Linear Type Theory for Resource Aware Certified Programming

### 2021-09-22 writing/draft.tex
writing

  \subsubsection{Substitution}
  Though the \textit{substitution} lemma is generally considered a boring and bureaucratic theorem, it is surprisingly hard to design linear typed languages where the \textit{substitution} lemma is admissible. Much of this difficulty arises during the substitution of non-linear expressions. Perhaps the most famous example detailing the issues of substitution is due to Wadler\cite{substitute}. Since computation arises as a consequence of substitution, it is paramount to get it right.

@@

@@

  \begin{theorem}
    Partial Weakening. If $\Gamma \vdash m \stype{s} A$ then $\Gamma, x \utype B \vdash m \stype{s} A$ is derivable.
  \end{theorem}

>>>>

  \subsubsection{Substitution}
  Though the \textit{substitution} lemma is widely considered a boring and bureaucratic theorem, it is surprisingly hard to design linear typed languages where the \textit{substitution} lemma is admissible. Much of this difficulty arises during the substitution of non-linear expressions. Perhaps the most famous work detailing the issues of substitution is due to Wadler\cite{substitute}. Since computation arises as a consequence of substitution, it is paramount to get it right.

  Generally, the application rule looks similar to the following for languages with linear types.
  \begin{mathpar}
    \inferrule
    { \Gamma \vdash m : A \multimap B \\ 
      \Delta \vdash n : A }
    { \Gamma, \Delta \vdash m\ n : B } 
  \end{mathpar}

  If type $A$ is non-linear, then $n$ ought to be used freely. However, it is possible for $n$ to contain linear variables in $\Delta$. If $m$ is a lambda abstraction $\lambda x.m'$ where $x$ occurs multiple times within $m'$, substitution of $n$ for $x$ will cause duplication of linear variable. A common approach for solving this issue is to wrap non-linear values in an explicit modality, unpacking the internal value only when needed\cite{substitute,neel15}. Another is to outright ban non-linear expressions from containing linear variables\cite{llf,luo,qtt}.

  Our call-by-value semantics neatly solves the substitution problem without imposing any of the modifications mentioned previously to the typing of application. The crucial realization is the following \textit{value soundness} lemma: non-linear \textit{values} contain no linear variables.

@@

  From the \textit{value soundness} lemma, the standard rule for application is admissible. Intuitively, a single copy of each linear resource is used to create a single non-linear value. This value can then be freely duplicated without needing the original resources to generate fresh copies.

@@

  \subsubsection{Type Soundness}
  The following theorems are a direct result of the \textit{substitution} lemma and various canonical-form lemmas.

  \begin{theorem}
    Non-linear Weakening. If $\Gamma \vdash m \stype{s} A$ then $\Gamma, x \utype B \vdash m \stype{s} A$ is derivable.
  \end{theorem}
